---
title: "Elementary Statistics using Excel"
author: 
- "Kory Illenye"
- "MAT 152 - Statistical Methods I"
- "Chapter 5.1,5.2"

date: '`r format(Sys.time(), "%b %d, %Y")`'
output: ioslides_presentation
---

<style>

body {
  background-color: #000029;
}

body p {
  color: blue;
  font-size: 15px;
}

body hgroup h2 {
  color: black;
}
.footer {
    color: White;
    background: Dodgerblue;
    position: fixed;
    top: 88%;
    text-align: center;
    width: 100%;
}

ul {
  display: block;
  font-size: 20px;
  margin-top: 0;
  margin-bottom: 0;
  margin-left: 0;
  margin-right: 0;
  font-weight: bold;
  color: blue;
}

ol {
  display: block;
  font-size: 20px;
  margin-top: 0;
  margin-bottom: 0;
  margin-left: 0;
  margin-right: 0;
  font-weight: bold;
  color: blue;
}

slides > slide.title-slide p {
  color: blue;
  font-size: 15px;
}



slides > slide.title-slide hgroup h1 {
  color: black;
  font-size: 40px;
  letter-spacing: 0;
}



slides > slide.{
  background: none !important;
  background-color: #444444 !important;
}

slides > slide.backdrop {
  background: white;
}

</style>

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = FALSE)
library(plotly)
library(dplyr)
library(qcc)
library(ggplot2)
```

## Chapter 5 Section 1 Objectives

<H3><font color="black">Probability Distributions</font></H3>
<br>


- Define random variable and probability distributions
- Determine When a potential probability distribution actually satisfies the necessary requirements.
- Given a probability distribution, compute the mean and standard deviation, then use those results to determine whether the results are significantly low or significantly high.

## Definitions

- **random variable** - a variable (generally $x$) that has a sngle numerical value, determined by chance, for each outcome of a procedure.
- **probability distribution** a description that gives the probability for each value of the random variable. It is often expressed in the format of a table, formula, or graph.
- **discrete randome variable** - a collection of values that is finite or countable. (Coin Flips)
- **continuous random variable** - Infinitely many values, and the collection is not countable. (Body Temp)
- **expected value** - the mean value of a dsicrete random variable. Often denoted as $E$.

## Requirements for Probability Distributions

1) There is a numerical random variable $x$ and its number values are associated with corresponding probability values.
3) $0 \leq P(x) \leq 1$ for every value of the random varaible $x$.

## Sum of two random Dice Rolls

| Sum of dice | Die roll Possibilities        | $P(x)$|Sum of dice    | Die roll Possibilities  | $P(x)$| 
|-------------|-------------------------------|-------|---------------|-------------------------|-------|
| 2           | (1,1)                         | 1/36  | 8             |(2,6)(3,5)(4,4)(5,3)(6,2)| 5/36  |
| 3           | (1,2)(2,1)                    | 2/36  | 9             |(3,6)(4,5)(5,4)(6,3)     | 4/36  |
| 4           | (1,3)(2,2)(3,1)               | 3/36  | 10            |(4,6)(5,5)(6,4)          | 3/36  |
| 5           | (1,4)(2,3)(3,2)(4,1)          | 4/36  | 11            |(5,6)(6,5)               | 2/36  |
| 6           | (1,5)(2,4)(3,3)(4,2)(5,1)     | 5/36  | 12            |(6,6)                    | 1/36  |
| 7           | (1,6)(2,5)(3,4)(4,3)(5,2)(6,1)| 6/36  |               |                         |       |

## Do the die rolls meet the criteria?

1) The variable $x$ is a numerical random variable and its values are associated with the previous table.
2) $\Sigma P(x) = \frac{1}{36} + \frac{2}{36} + \frac{3}{36} + \frac{4}{36} + \frac{5}{36} + \frac{6}{36} + \frac{5}{36} + \frac{4}{36} + \frac{3}{36} + \frac{2}{36} + \frac{1}{36} = 1$
3) Each value of $P(x)$ is between 0 and 1.

## Histogram

```{r}
x <- c(2,3,3,4,4,4,5,5,5,5,6,6,6,6,6,7,7,7,7,7,7,8,8,8,8,8,9,9,9,9,10,10,10,11,11,12)
#hist(x, freq = FALSE, xlab = "Sum of Two Dice", ylab = "Probability", main = "Histrogram of Sum of two dice", breaks = 11)
qplot(x,y=..density.., geom = "histogram", bins = 11,fill=I("blue"),col=I("black"))+ labs(x = "Sum of Two Dice", y = "Probability", title= "Histogram of the sum of two dice")
```

## Parameters of Probability Distribution

- $\mu = \Sigma [ x \cdot P(x)]$ also known as $E[x]$
- $\sigma^2 = \Sigma [x-\mu^2] \cdot P(x)$ 
- $E[x^2] = \Sigma [x^2 \cdot P(x)]$
- $\sigma^2 = E[x^2] - (E[x])^2$
- $\sigma = \sqrt{E[x^2] - (E[x])^2}$

## Significance

- Significantly Low - Falls below or equal to $\mu - 2 \cdot \sigma$
- Significantly High - Falls above or equal to $\mu + 2 \cdot \sigma$
- Not significant - Falls between $\mu - 2 \cdot \sigma$ and $\mu + 2 \cdot \sigma$.



## Chapter 5 Section 2 Objectives

<H3><font color="black">Binomial Probability Distributions</font></H3>
<br>

- Describe a binomial probability distribution and find the probability values for a binomial distribution.
- Compute the mean and standard deviation for a binomial distribution, then use those results to determine whether the results are significantly low or significantly high.

## Definitions

- **binomial probability distribution** - results from a procedure that meets the following:
    + The procedure has a fixed number of trials.
    + The trials must be independent.
    + Each trial must have all outcomes classified into exactly 2 categories. (Bernoulli)
    + The probability of remains the same on all trials.
    
## Binomial Probability Distributions

- $P(S) = p$, $p$ is the probability of success
- $P(F) = 1 - p = q$, $q$ is the probability of failure
- $n$ is a fixed number of trials
- $x$ is a specific number of success in $n$ trials.
- $P(x)$ is the probability of getting exactly $x$ successes in $n$ trials

## Binomial Formula

$$\Huge P(x) = \binom{n}{x} \cdot p^x \cdot q^{n-x}$$ 
$$\Large \textrm{for }x = 1,2,3,....,n$$

## Using Excel

- BINOM.DIST
    + =BINOM.DIST(x,n,p,0)
    + the last 0 means you want the the probability of $x$ of success. 
    + Change the last 0 to a 1 and it will give you the cummalitve proability to $x$ success. So the probability of getting x and fewer successes.
    
## Mean, Variance and Standard deviation of Binomial Distribution

- $\mu = np$
- $\sigma^2 = npq$
- $\sigma = \sqrt{npq}$