---
title: "Elementary Statistics using Excel"
author: 
- "Kory Illenye"
- "MAT 152 - Statistical Methods I"
- "Chapter 6.1,6.2,6.3"

date: '`r format(Sys.time(), "%b %d, %Y")`'
output: ioslides_presentation
---

<style>

body {
  background-color: #000029;
}

body p {
  color: blue;
  font-size: 15px;
}

body hgroup h2 {
  color: black;
}
.footer {
    color: White;
    background: Dodgerblue;
    position: fixed;
    top: 88%;
    text-align: center;
    width: 100%;
}

ul {
  display: block;
  font-size: 20px;
  margin-top: 0;
  margin-bottom: 0;
  margin-left: 0;
  margin-right: 0;
  font-weight: bold;
  color: blue;
}

ol {
  display: block;
  font-size: 20px;
  margin-top: 0;
  margin-bottom: 0;
  margin-left: 0;
  margin-right: 0;
  font-weight: bold;
  color: blue;
}

slides > slide.title-slide p {
  color: blue;
  font-size: 15px;
}



slides > slide.title-slide hgroup h1 {
  color: black;
  font-size: 40px;
  letter-spacing: 0;
}



slides > slide.{
  background: none !important;
  background-color: #444444 !important;
}

slides > slide.backdrop {
  background: white;
}

</style>

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = FALSE)
library(plotly)
library(dplyr)
library(qcc)
library(ggplot2)
```

## Chapter 6 Section 1 Objectives

<H3><font color="black">The Standard Normal Distribution</font></H3>
<br>


- Describe the characteristics of a standard normal distribution
- Find the probability of some range of z-values in a standard normal distribution
- Find z scores to regions under the curve representing a standard normal distribution

## Definitions 6.1

- **Normal distribution** - a continuous random variable has a distribution with a graph that is symmetric and bell-shaped.
- **Uniform distribution** - a continuous random variable has an even spread over a range of possibilities. 
- **Density Curve** - A graph of any continuous probability distribution. It must have a total area under the curve is exactly 1.
- **Standard Normal Distribution** - a normal distribution with the parameters of $\mu = 0$ and $\sigma = 1$. The total area under the curve is equal to 1.
- **Critical Value** - a z-score that is on the borderline that separates the z-scores that are significantly high or low

## Normal Distribution

$$ y = \frac{e^{-\frac{1}{2}(\frac{x-\mu}{\sigma})^2}}{\sigma\sqrt{2\pi}}$$
```{r}
x <- seq(-4, 4, length=100)
hx <- dnorm(x)
plot(x, hx, type="l", lty=2, xlab="x value",
  ylab="Density", main="Standard Normal Distribution", col = "blue")
```


## Uniform Distribution

- Whats the area under the curve?

```{r}
curve(dunif(x, min = 2, max = 6), 0, 8, ylim = c(0,0.5), ylab = "f( x )" , main = "Uniform Density f( x )", lty = 2, col = "blue")

```

## Finding Probabilities when given a *z*-score

- Table A-2 in your book.
- Each value represents the area under the curve to the left of the *z*-score
- Using Excel
    + use **NORM.DIST**
    + =**NORM.DIST**(x, mean, standard_deviation, cumulative)
    + x is the *z*-score
    + mean (for standard normal the mean is 0)
    + standard deviation (for standard normal the standard deviation is 1)
    + cumulative tells if you wan the area to the left up to the *z*-score1
    
## An Example

- A randomly selected adult undergoes a bone density test. Find the probability that this person has a bone density test score of less than 1.73. The test scores for bone density have a mean of 0 and a standard deviation of 1.
    - What do you get from Table A-2?
    - What do you get from excel?
    - Are they similar? 
    - What about a score greater than 1.73?

## Finding the area between two values

```{r}
mean=0; sd=1
lb=-1.23; ub=1.73

x <- seq(-4,4,length=100)*sd + mean
hx <- dnorm(x,mean,sd)

plot(x, hx, type="n", xlab="IQ Values", ylab="",
  main="Normal Distribution", axes=FALSE)

i <- x >= lb & x <= ub
lines(x, hx)
polygon(c(lb,x[i],ub), c(0,hx[i],0), col="blue") 

area <- pnorm(ub, mean, sd) - pnorm(lb, mean, sd)
result <- paste("P(",lb,"< IQ <",ub,") =",
   signif(area, digits=3))
mtext(result,3)
axis(1, at=seq(-3, 3, 1), pos=0)
```

## Finding a *z*-score from known area

- Use table A-2
- Use Excel NORM.INV
    + =NORM.INV(p, mean, standard deviation)
    + p is the probability area
    + mean use 0 if you are looking for a *z*-score
    + standard deviation use 1 if you are looking for a *z*-score

## Chapter 6 Section 2 Objectives

<H3><font color="black">Real Applications of Normal Distributions</font></H3>
<br>

- Develop the ability to describe a normal distribution
- Find the probability of some range of values in a normal distribution
- Find x scores corresponding to regions under the curve representing a normal distribution

## *z*-score formula

$$ z = \frac{x-\mu}{\sigma}$$

- The *z*-score formula can allow you to use the table A-2 on non-standard normal distributions.
- The mean body temperature is 98.6 degrees Fahrenheit with a standard deviation of 0.73 degree Fahrenheit according to the University of Wisconson - madison. 
- What is the probability that a person has a body temperature of 97.2 degrees Fahrenheit?
    - Table A-2?
    - Excel?
    
## Airforce Height of pilots

- In the airfore pilots are required to be 64 to 77 in. and women have an average height of 63.7 in. with a standard deviation of 2.9 in.. What percent of women meet the height requirement?

```{r}
mean=63.7; sd=2.9
lb=64; ub=77

x <- seq(-4,4,length=100)*sd + mean
hx <- dnorm(x,mean,sd)

plot(x, hx, type="n", xlab="Height", ylab="",
  main="Normal Distribution", axes=FALSE)

i <- x >= lb & x <= ub
lines(x, hx)
abline(v = mean, lty = 2, col ="red")
polygon(c(lb,x[i],ub), c(0,hx[i],0), col="blue") 

area <- pnorm(ub, mean, sd) - pnorm(lb, mean, sd)
result <- paste("P(",lb,"< Height <",ub,")")
mtext(result,3)
axis(1, at=seq(50, 77, 1), pos=0)
```



## Chapter 6 Section 3 Objectives

<H3><font color="black">Sampling Distributions and Estimators</font></H3>
<br>

- Develop the ability to describe a sampling distribution of a statistic
- Determine whether a statistic serves as a good estimator of the corresponding population parameter

## Definitions 6.3

- **Sampling distribution of a statistic** - The distribution o fall values of the statistic when all possible samples of the same size *n* are taken from a the same population.
- **Sampling distribution of a sample proportion** - The distribution of sample proportions with all samples having the same sample size *n* taken from the same population.
- **Sampling distribution of the sample mean** - The distribution of all possible sample means with all samples having the same sample size *n* taken from the same population.
- **Sampling distribution of the sample variance** - The distribution of all possible sample variances with all samples having the same sample size *n* taken from the same population.
- **Estimator** - Statistic used to estimate a population parameter
- **Unbiased estimator** - Statistic that targets the value of the corresponding population parameter in the sense that the sampling distribution of the statistic has a mean that is equal to the correspond population parameter.

## Behavior of sample proportions

- Sample proportions tend to be normally dsitributed
- The mean of sample proportions is the same as the population proportion

## Behavior of sample means

- Sample means tend to be normally distributed
- Sample means target the population mean

## Behavior of sample variances

- Sample variances tend to be skewed right
- Sample variances target teh population variance

## Unbiased vs. Biased estimators


| Unbiased                          | Biased                                  | 
|-----------------------------------|-----------------------------------------|
| Proportion ($\hat{p}$)            | Median                                  | 
| Mean ($\bar{x}$)                  | Range                                   |
| Variance ($s^2$)                  | Standard Deviation                      |
| *Target the population parmeter*  | *Does not target population parameter*  |

