---
title: "Elementary Statistics using Excel"
author: 
- "Kory Illenye"
- "MAT 152 - Statistical Methods I"
- "Chapter 7.2,7.3"

date: '`r format(Sys.time(), "%b %d, %Y")`'
output: ioslides_presentation
---

<style>

body {
  background-color: #000029;
}

body p {
  color: blue;
  font-size: 15px;
}

body hgroup h2 {
  color: black;
}
.footer {
    color: White;
    background: Dodgerblue;
    position: fixed;
    top: 88%;
    text-align: center;
    width: 100%;
}

ul {
  display: block;
  font-size: 20px;
  margin-top: 0;
  margin-bottom: 0;
  margin-left: 0;
  margin-right: 0;
  font-weight: bold;
  color: blue;
}

ol {
  display: block;
  font-size: 20px;
  margin-top: 0;
  margin-bottom: 0;
  margin-left: 0;
  margin-right: 0;
  font-weight: bold;
  color: blue;
}

slides > slide.title-slide p {
  color: blue;
  font-size: 15px;
}



slides > slide.title-slide hgroup h1 {
  color: black;
  font-size: 40px;
  letter-spacing: 0;
}



slides > slide.{
  background: none !important;
  background-color: #444444 !important;
}

slides > slide.backdrop {
  background: white;
}

</style>

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = FALSE)
library(plotly)
library(dplyr)
library(qcc)
library(ggplot2)
```

## Chapter 7 Section 2 Objectives

<H3><font color="black">Estimating a Population Mean</font></H3>
<br>


- Construct a confidence interval estimate of a population mean, and be able to interpret such confidence interval.
- Determine the sample size necessary to estimate a population mean


## Estimating a population mean when $\sigma$ is not known

- It is rare to estimate a a population mean ($\mu$) and also know the standard deviation ($\sigma$).
- The sample mean ($\bar{x}$) is the best point estimate of $\mu$.
    + $\bar{x}$ gives no clue to how accurate it is at estimating $\mu$.
    + A confidence interval (CI) must be constructed to show accuracy.

## Requirements of a confidence interval for estimating $\mu$ with $\sigma$ not known

- The sample is a simple random sample.
- The sample meets one of the following:
    + The population is normally distributed.
    + The sample size ($n$) is greater than 30.
    
## Margin of Error

$$\Huge E = t_{\frac{\alpha}{2}}\cdot \frac{s}{\sqrt{n}}$$

- degrees of freedom (df) is $n-1$.
- $\alpha$ is 1 - confidence level.
- $n$ is the sample size
- $s$ is the sample standard deviation

## Confidence interval and the students t distribution

$$\Huge \bar{x} \pm E$$

```{r}
x <- seq(-3, 3, length=1000)
hx <- dnorm(x)

degf <- c(1, 2, 5, 10)
colors <- c("red", "blue", "darkgreen", "gold", "black")
labels <- c("df=1", "df=2", "df=5", "df=10", "normal")

plot(x, hx, type="l", lty=1, xlab="x value",
     ylab="Density", main="Comparison of t Distributions")

for (i in 1:4){

  if (i == 1) {
    lines(x, dt(x,degf[i]), lwd=3, col=colors[i])
  } 
  else
  {
    lines(x, dt(x,degf[i]), lwd=1, col=colors[i])
  }
}

legend("topleft", inset=.05, title="Distributions",
       labels, lwd=1, lty=c(1, 1, 1, 1, 1), col=colors)
```

## Using excel to find $t_{\frac{\alpha}{2}}

- $t_{\frac{\alpha}{2}}$ =  T.INV.T2($alpha$,df)
- remember df = $n-1$ and $\alpha$ = 1 - confidence level
- Given 15 data points:
    + 11,12,13,14,15,16,17,18,19,20,21,22,23,24,25
- Find the 95% confidence interval of the mean

## Interpreting the Confidence interval

- We are 95\% confident that the interval of lower limit to upper limit contains the true population mean of $\mu$.

## Determining sample size needed

$$\Huge n = (\frac{z_{\frac{\alpha}{2}}\cdot \sigma}{E})^2 $$

- $E$ is the desired amount of error
- $\sigma$ population standard deviation. Usually unknown
    + Range rule of thumb $\frac{\textrm{range}}{4}$
    + Use sample standard deviation
    + Use results from an earlier study

## Confidence interval when $\sigma$ is known

$$\Huge E = z_\frac{\alpha}{2}\cdot\frac{\sigma}{\sqrt{n}}$$

- Use excel CONFIDENCE.NORM
    + $E$ = CONFIDENCE.NORM($\alpha$,$\sigma$, $n$)
- CI = $\bar{x} \pm E$





## Chapter 7 Section 3 Objectives

<H3><font color="black">Estimating a Population Standard Deviation or Variance</font></H3>
<br>


- Develop the ability to construct a confidence interval estimate of a population standard deviation or variance, and be able to interpret such confidence interval estimates.


## Chi-Square Method of estimating variance

- $s^2$ is the best point estimate for $\sigma^2$

$$\Huge \chi^2 = \frac{(n-1)\cdot s^2}{\sigma^2}$$

```{r}
colors = c("black", "red")
labels = c("df 4", "df 5")
ggplot(data.frame(x = c(0, 20)), aes(x = x)) +
     stat_function(fun = dchisq, args = list(df = 4)) + stat_function(fun = dchisq, args = list(df = 5), col = "red")+ stat_function(fun = dchisq, args = list(df = 10), col = "green")

```

## Using excel to find a confidence interval

- Left and right are different so must be done separately
- CHISQ.INV ($\chi^2_L$) and CHISQ.INV.RT ($\chi^2_R$)


$$\Huge (\sigma^2 - \frac{(n-1)\cdots^2}{\chi^2_R}, \sigma^2 + \frac{(n-1)\cdots^2}{\chi^2_L})$$

$$\Huge (\sigma - \sqrt{\frac{(n-1)\cdots^2}{\chi^2_R}}, \sigma + \sqrt{\frac{(n-1)\cdots^2}{\chi^2_L}})$$


## For determining sample size of $\chi^2$ distribution

- This is very complex due to the distribution not being normal
- Refer to table 7-2 on page 365 of your book.