---
title: "Graphing Rational Functions"
author: "Kory Illenye"
date: "6/27/2018"
output: ioslides_presentation
---

<style>

body {
  background-color: Dodgerblue;
}

body p {
  color: blue;
}

body hgroup h2 {
  color: black;
}
.footer {
    color: White;
    background: Dodgerblue;
    position: fixed;
    top: 92%;
    text-align: center;
    width: 100%;
}

ul {
  display: block;
  font-size: 20px;
  margin-top: 0;
  margin-bottom: 0;
  margin-left: 0;
  margin-right: 0;
  font-weight: bold;
  color: blue;
}



slides > slide.title-slide p {
  color: blue;
  font-size: 15px;
}



slides > slide.title-slide hgroup h1 {
  color: black;
  font-size: 40px;
  letter-spacing: 0;
}



slides > slide.{
  background: none !important;
  background-color: #444444 !important;
}

slides > slide.backdrop {
  background: white;
}
</style>

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = FALSE, prompt = FALSE, comment = NA, warning = FALSE)
library(matrixcalc)
```

<div class="footer">Graphing Rational Functions by Kory Illenye</div>
## Rational Expressions and Functions

**Blitzer definitions:**

A **Rational Expression** consists of a polynomial divided by a nonzero polynomial (denominator cannot be equal to 0). 

A **Rational Function** is a function defined by a formula that is a rational expression. For Example:

$$ f(x) = \frac{x+3}{x+6}$$

## Domain of Rational Functions

The domain of a rational function is all real numbers except those that make the denominator equal to zero.

Example:
$$f(x) = \frac{x + 3}{x^2-9}$$
Set the denominator equal to zero and solve.
$$x^2 -9 = 0$$
Using the square root property we know that $x =\pm 3$ solves this equation. So $\pm 3$ is not in the domain.

Interval notation: $D = (-\infty,-3)\cup(-3,3)\cup(3,\infty)$

## Vertical Asymptotes and holes

A **vertical asymptote** is a vertical line that the graph of a function approaches, but does not touch.

Any value that would make the denominator equal to zero is either a hole or a vertical asymptote.

Lets look at our previous function:
$$f(x) = \frac{x + 3}{(x+3)(x-3)}$$
at $x = 3$ we have a vertical asymptote and at $x=-3$ we have a hole. Why? can you figure it out?

## Horizontal Asymptotes

a **horizontal asymptote** is a horizontal line that the graph of a function approaches as $x$ gets very large or very small.  The graph of a function may touch/cross its horizontal asymptote in multiple places.

lets define a general rational function:
$$f(x) = \frac{a_mx^m + a_{m-1}x^{m-1}+a_{m-2}x^{m-2}+...+a_0}{b_nx^n + b_{n-1}x^{n-1}+b_{n-2}x^{n-2}+...+b_0}$$
$m$ represents the degree of the polynomial in the numerator and $n$ denotes the degree of the polynomial in the denominator. 

$a$ represents the coefficients of the polynomial in numerator and $b$ denotes the coefficients of the polynomial in the denominator.

## Horizontal Asymptote (continued)

- if $m < n$ then there is a horizontal asymptote at $y=0$.
- if $m = n$ then there is a horizontal asymptote at $y = \frac{a_m}{b_n}$.
- if $m > n$ then no horizontal asymptote exists.

## Graphing a Rational Function

- step 1: factor both the numerator and denominator completely.
- step 2: state the domain.
- step 3: simplify if able.
- step 4: identify vertical asymptotes and holes.
- step 5: idtenify horizontal asymptors.
- step 6: find $x$ and $y$ intercepts.
- step 7: create a behavior table for points near all vertical asymptotes for both sides of the asymptote.
- step 8: plot graph

## Graphing a Rational Function Example

$$f(x) = \frac{3x+9}{x^2-9}$$

- step 1: $f(x) = \frac{3(x+3)}{(x-3)(x+3)}$
- step 2: domain is all real numbers except 3 and -3. $D = (-\infty,-3)\cup(-3,3)\cup(3,\infty)$
- step 3: $\frac{3}{x-3}$
- step 4: vertical asymptote at $x = 3$ and a hole at $x = -3$
- step 5: horizontal asymptote at $y=0$
- step 6: no $x$ intercept and a $y$ intercept at $(0,-1)$
- step 7: $f(2.9) = -20.16949$ and $f(3.1) = 19.83607$

## plot
```{r}
curve(x+1, -20,20, col= "white")
curve((3*x +9)/(x*x -9), -20, -3.2, add= TRUE)
curve((3*x +9)/(x*x -9), -2.8, 2.9999, add= TRUE)
curve((3*x +9)/(x*x -9), 3.00001,20, add= TRUE)
points(-3,-1/2)
points(3.1,19.84, pch=20)
points(2.9,-20.17, pch=20)
points(0,-1, pch=20)
abline(v=0, col="black", lwd=1)
abline(h=0, col="black", lwd=1)
abline(v=3, col="blue", lty=2, lwd=2)
abline(h=0, col="blue", lty=2, lwd =2)
```

## Questions

\center Questions?

## References 

Blitzer, Robert, *"Algebra and Trigonometry 6th edition"*, 2018.
